\documentclass[letterpaper,10pt]{article}
\usepackage[margin=0.72in]{geometry}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{float}
\usepackage{graphicx}
\usepackage{url}
\setlength{\parindent}{0pt}
\setlength{\parskip}{0.45em}
\renewcommand{\arraystretch}{1.08}

\title{Technical Appendix for Anonymous Review}
\author{}
\date{}

\begin{document}
\maketitle
\small

\section{Resource and Construction Details}

CliniCause provides two dataset-specific, estimator-ready causal-analysis
resources: MIMIC-III (26,845 analysis records) and PhysioNet 2012 (7,993).
The analysis unit is a source time-series record or ICU stay, rather than a
unique person.  At each source boundary, accepted identifiers are normalized to
the canonical \texttt{ts\_id} key; long irregular events retain minute,
variable, and value, while a record-level table carries in-hospital mortality
and available covariates.  The two resources share an interface but are never
pooled: their variable vocabularies, proxy ontologies, measurement processes,
and DAG assumptions remain source specific.

\begin{table}[H]
\centering\small
\caption{Admitted original-cohort exposure sets and source-specific semantics.}
\label{tab:supp-resources}
\begin{tabular}{p{0.19\linewidth}p{0.35\linewidth}p{0.37\linewidth}}
\toprule
Resource & Admitted exposures & Source-specific proxy boundary\\
\midrule
MIMIC-III & Cardiac strain; inflammation/sepsis; hepatic/coagulation dysfunction; renal dysfunction; global severity; metabolic derangement; neurologic dysfunction; respiratory failure; shock. & Hepatic and coagulation evidence are represented by a combined proxy.  Chronic-burden fields are retained as baseline constructs rather than admitted exposures.\\
PhysioNet 2012 & Cardiac injury/strain; coagulation/hematologic dysfunction; global severity; hepatic dysfunction; inflammation/sepsis burden; metabolic derangement; neurologic dysfunction; renal dysfunction; respiratory failure; shock. & Hepatic and coagulation/hematologic evidence are represented separately.  Chronic baseline constructs remain outside the admitted exposure set.\\
\bottomrule
\end{tabular}
\end{table}

At design time, structured LLM proposals covered proxy ontologies, rule
families, missingness considerations, and dataset-specific DAGs.  Project
selection encoded chosen proposals as deterministic proxy-rule and DAG source
artifacts.  The LLM did not receive runtime patient rows and was not a
patient-level estimator.  At construction time, each source is preprocessed in
its own lane; deterministic rule labels and four temporal models (STraTS, GRU,
GRU-D, and TCN) produce aligned annotation exports.  Each export is normalized
to probability and binary fields; one deterministic rule source and four model
sources are then combined by a fixed deterministic aggregation.  The final
resource joins the aggregate exposure layer to outcome, observed covariates,
DAG/adjustment metadata, and provenance records.

Validation is layered: schema handoffs check required columns, order, value
ranges, binary/probability consistency, and expected output sets; cohort
handoffs reject conflicting IDs, missing/extra records, and silent shrinkage;
prediction exports and aggregation enforce complete aligned voters; and
provenance artifacts bind available hashes, manifests, configurations, and
receipts.  Archived-result evidence establishes checked aggregate outputs, but
the exact historical producing revisions, configurations, and
checkpoint-to-export lineage remain incomplete.  Current repaired contracts are
documented separately and are not claimed to have produced the archived results.

\section{Complete Predictive Evaluation}

The predictive task is multi-label estimation of each dataset's deterministic
proxy vector.  Test loss is sample-mean multi-label binary cross-entropy.
AUROC, AUPRC, and minRP are calculated per target then macro-averaged after
excluding target columns with a single observed evaluation class; minRP is the
maximum over thresholds of the smaller of precision and recall.  The table
reports selected archived held-out point metrics.  It does not supply repeated
run uncertainty, significance testing, split-ID lineage, or a verified
checkpoint-to-export mapping.

\input{generated/predictive_table.tex}

\section{Complete Original-Cohort Estimator Matrix}

\input{generated/original_estimator_table.tex}

\section{Outcome-Downsampling Comparison}

All mortality-positive records are retained while mortality-negative records are
sampled under the implemented configuration.  This produces a different
analyzed population and outcome prevalence; therefore, the table uses direction
as a robustness comparison and does not treat downsampled magnitudes as
transportable estimates for the original populations.  The generator verifies
all 57 matched original/downsampled estimator rows and exactly two sign changes.

\input{generated/downsampling_table.tex}

\section{Matching Availability and Support}

Matching is a descriptive, support-oriented, greedy one-to-one comparison over
the archived binary matching representation.  Successful rows report the
descriptive matched-pair outcome difference.  A failure indicates that no usable
binary matching columns remained after preprocessing; it is not an effect of
zero.

\input{generated/matching_table.tex}

\section{Sensitivity and Permutation Coverage}

\input{generated/diagnostic_coverage_table.tex}

\section{Provenance and Reproducibility Boundaries}

Checked archived artifacts support the resource counts, selected predictive
metrics, original and downsampled estimator summaries, matching availability,
and the recorded diagnostic status classes.  They do not establish an exact
historical clean-checkout rerun.  The known gaps include producing source
revisions, numbered causal configurations, predictive split IDs,
checkpoint-to-export linkage, raw/processed-data manifests, producing
environment/hardware capture, and CausalPFN package/version alignment.

Current source inspection documents stricter runtime validation gates for
schemas, canonical identifiers, voters, and artifact reuse.  Those current
contracts must be tested and release-validated separately; this appendix makes
no current test-pass claim.  Any eventual anonymous package must keep
source-data access restricted to authorized MIMIC-III and PhysioNet users,
include only licensing-permitted derived artifacts, and undergo a final identity
and metadata audit.  No patient-level values are included in this appendix.

\end{document}
