\chapter*{Abbreviations}
\addcontentsline{toc}{chapter}{Abbreviations}

\begin{longtable}{@{}p{0.20\textwidth}p{0.72\textwidth}@{}}
\toprule
Abbreviation & Meaning in this thesis \\
\midrule
\endfirsthead
\toprule
Abbreviation & Meaning in this thesis \\
\midrule
\endhead
\bottomrule
\endfoot
ICU & Intensive Care Unit. \\
EHR & Electronic Health Record. \\
DAG & Directed Acyclic Graph. \\
LLM & Large Language Model; used in this thesis for design-time knowledge elicitation, not runtime patient-level inference. \\
STraTS & Self-Supervised Transformer for Sparse and Irregularly Sampled Multivariate Clinical Time Series. \\
GRU & Gated Recurrent Unit. \\
GRU-D & Gated Recurrent Unit with Decay. \\
TCN & Temporal Convolutional Network. \\
SAnD & Simply Attend and Diagnose. \\
DML & Double Machine Learning. \\
CATE & Conditional Average Treatment Effect; reported here only as a model-estimated quantity under the stated assumptions. \\
HTE & Heterogeneous Treatment Effect. \\
AUROC & Area Under the Receiver Operating Characteristic Curve. \\
AUPRC & Area Under the Precision--Recall Curve. \\
minRP & Minimum of precision and recall. \\
SOFA & Sequential Organ Failure Assessment. \\
ARDS & Acute Respiratory Distress Syndrome. \\
CausalPFN & Exploratory model name used in the project; its primary methodological source and expansion are unresolved. \\
\end{longtable}

\clearpage
\chapter*{Notation and Symbols}
\addcontentsline{toc}{chapter}{Notation and Symbols}

\begin{longtable}{@{}p{0.25\textwidth}p{0.67\textwidth}@{}}
\toprule
Symbol & Meaning in this thesis \\
\midrule
\endfirsthead
\toprule
Symbol & Meaning in this thesis \\
\midrule
\endhead
\bottomrule
\endfoot
\(i\) & Index of a normalized analysis record. \\
\(n_i\) & Number of observed measurement events for record \(i\). \\
\(t_{ij}\) & Elapsed time in minutes for event \(j\) of record \(i\). \\
\(v_{ij}\) & Observed variable name for event \(j\) of record \(i\). \\
\(x_{ij}\) & Numeric value recorded for event \(j\) of record \(i\). \\
\(\mathcal{V}\) & Set of observed measurement-variable names. \\
\(\mathcal{O}_i\) & Long-format measurement-event object for record \(i\). \\
\(L^{rule}_{ik}\) & Rule-derived binary proxy label for record \(i\) and proxy state \(k\). \\
\(\hat p^{(m)}_{ik}\) & Probability for proxy state \(k\) exported by model \(m\) for record \(i\). \\
\(\hat L^{(m)}_{ik}\) & Binary predicted proxy label corresponding to \(\hat p^{(m)}_{ik}\). \\
\(\widetilde L_{ik}\) & Aggregated predicted proxy-label representation; in the archived final causal runs, deterministic voting combined one rule-derived source with four aligned binary model-prediction sources. \\
\(A_i\) & Selected binary proxy-state exposure for record \(i\). \\
\(Y_i\) & In-hospital mortality outcome for record \(i\). \\
\(W_i\) & Observed adjustment variables selected from the project-specified DAG and available columns. \\
\(X_i\) & Effect-modifier features used to describe heterogeneity. \\
\(Y_i(a)\) & Conceptual potential outcome under exposure value \(a\); its interpretation requires the assumptions stated in the methods. \\
\end{longtable}
