\chapter{Reproducibility and Evidence Boundary}
\label{app:reproducibility-evidence-boundary}

\section{Current reproducibility interface}

CliniCause now exposes a unified execution interface that can
reconstruct causal-analysis resources from appropriately obtained raw inputs
and execute the integrated predictive and causal workflow. The repository
provides one canonical shell launcher, one project-wide dependency entry point,
\texttt{requirements.txt}, dataset-isolated run
directories, and explicit validation, cohort, artifact, receipt, and provenance
contracts. This makes the current implementation and its reconstruction path
substantially more reproducible without redistributing source ICU records.

\section{Repository setup}

On a fresh machine, clone the repository normally. The checkout contains the
complete pipeline, including the \texttt{STraTS/} and
\texttt{causal-irregular-time-series/} component directories:

\begin{quote}\small
\begin{verbatim}
git clone https://github.com/KenziVisor/CliniCause.git
cd CliniCause
\end{verbatim}
\end{quote}

The first component contains the temporal-model implementation; the second
contains preprocessing, proxy construction, graph, and causal-analysis code.
No separate repository initialization step is required.

\section{Python 3.10 environment and dependencies}

Create an isolated Conda environment, activate it, upgrade its packaging tool,
and install the complete project through the root requirements file:

\begin{quote}\small
\begin{verbatim}
conda create -n clinicause python=3.10 -y
conda activate clinicause
python -m pip install --upgrade pip
python -m pip install -r requirements.txt
\end{verbatim}
\end{quote}

The root file declares the parent orchestration dependencies and recursively
includes the independently maintained requirement files of both component
directories.
It is the single supported installation interface for the current checkout;
it is not an environment lock or proof of the exact software environment used
to produce every archived thesis result.

\section{Source-data access}

MIMIC-III and the PhysioNet/Computing in Cardiology Challenge 2012 records are
not redistributed with CliniCause. Users must obtain their own copies from
PhysioNet and follow the access and license terms shown on the respective
project pages. MIMIC-III is a restricted resource: access requires PhysioNet
credentialing, the currently specified human-subjects training, and acceptance
of its data-use agreement. The Challenge 2012 files are currently presented
by PhysioNet as open access under their specified license; users must still
comply with that license and verify the current project-page terms. The
pipeline neither automates nor bypasses these processes, and restricted source
records must remain under the authorized user's control.

\section{Default and external data locations}

After obtaining and extracting the sources in a form consistent with their
official distributions, place each raw dataset root at the launcher default:

\begin{quote}\small
\begin{verbatim}
data/physionet2012
data/mimiciii
\end{verbatim}
\end{quote}

These paths denote the raw dataset roots consumed directly by the two
source-specific preprocessors; they are not locations for CliniCause-derived
outputs. To keep protected data outside the repository, supply absolute paths:

\begin{quote}\small
\begin{verbatim}
export PHYSIONET_RAW_DATA_PATH=/absolute/path/to/physionet2012
export MIMIC_RAW_DATA_PATH=/absolute/path/to/mimiciii
\end{verbatim}
\end{quote}

The launcher accepts absolute paths. It resolves relative values, including
the defaults, from the CliniCause repository root rather than from the caller's
current directory.

\section{Constructing the reusable datasets}

From the repository root and the activated environment, run:

\begin{quote}\small
\begin{verbatim}
STAGES=dataset-extraction bash run_clinicause.sh
\end{verbatim}
\end{quote}

The default dataset selector is \texttt{both}. This preset executes the
implemented preprocessing, deterministic tagging, tree generation, STraTS input
preparation and model stages, prediction collection and normalization, then
terminates the causal program at graph construction and deterministic majority
vote. Its terminal resource artifacts are the dataset-specific causal graph
and the aggregated latent-tag table, accompanied by a run summary. Thus the
preset reconstructs the reusable resource through the implemented aggregation
boundary; it does not run the later causal estimators.

By default, output is isolated below \path{runs/<run-id>/}, in separate
\texttt{physionet/} and \texttt{mimic/} children, with coordinator metadata
under the same run root. Dataset manifests, resolved configurations, stage logs, provenance
sidecars, cohort fingerprints, artifact hashes, and per-stage receipts record
and validate the current execution. A generated run identifier is used unless
\texttt{RUN\_ID} is set.

External raw-data paths can be applied to this command without copying the
records into the checkout:

\begin{quote}\small
\begin{verbatim}
PHYSIONET_RAW_DATA_PATH=/absolute/path/to/physionet2012 \
MIMIC_RAW_DATA_PATH=/absolute/path/to/mimiciii \
STAGES=dataset-extraction \
bash run_clinicause.sh
\end{verbatim}
\end{quote}

\section{Running the complete pipeline}

The canonical complete run is:

\begin{quote}\small
\begin{verbatim}
STAGES=all bash run_clinicause.sh
\end{verbatim}
\end{quote}

This retains the full integrated downstream predictive and causal workflow
rather than stopping at the dataset-construction boundary. It is
computationally heavier, and the temporal-model stages may require suitable
accelerator resources. Optional launcher controls include
\texttt{DATASET=physionet} or \texttt{DATASET=mimic} for a single source,
\texttt{OUTPUT\_ROOT}, \texttt{RUN\_ID}, the two raw-data path variables,
\texttt{STRATS\_MAX\_CONCURRENT}, and \texttt{PYTHON\_BIN}; the two commands
above remain the canonical interfaces.

\section{Reproducibility boundary and limitations}

Three claims must remain distinct. First, the current implementation is
reproducible at the level of repository composition, one dependency
interface, explicit launcher controls, dataset-isolated outputs, and validated
artifact contracts. Second, an authorized user can reconstruct the derived
resources from appropriately obtained raw inputs without CliniCause
redistributing those inputs. Third, neither achievement is automatically an
exact historical reproduction of the archived thesis results. The available
archive and compact reproducibility package provide checked result tables,
manifests, checksums, and partial run lineage, but do not recover every producing
revision, numbered configuration, split, checkpoint-to-export link, raw/processed
manifest, or hardware and software detail. Runtime can also vary across
systems. Finally, software reproducibility does not establish clinical
construct validity or causal identification; those remain separate scientific
and governance questions.
