\chapter{Introduction}
\label{chap:introduction}

\section{Motivation: Irregular ICU Time-Series Analysis}
\label{sec:introduction:motivation-irregular-icu-time-series-analysis}

Intensive-care medicine produces a particularly demanding form of longitudinal data.  Electronic health records contain laboratory values, vital signs, treatments, administrative fields, and outcomes, but these quantities are neither observed on a common clock nor collected with a common purpose.  Measurements occur at nonuniform times; variables are sampled at markedly different frequencies; many values are absent at most time points; and the intensity of measurement may itself reflect clinical concern, workflow, or resource use.  The resulting records are high dimensional and heterogeneous across patients, care trajectories, and institutions.  Irregular medical time-series methods therefore face more than a standard sequence-modeling inconvenience: they must represent the timing, absence, and changing availability of observations while preserving a defensible analytical link to the question being asked \cite{sun_2026_review_irregular_medical_timeseries,lipton_kale_wetzel_2016_missingness_rnns}.

The conventional convenience of a fixed, complete measurement matrix can conceal these properties.  Regular resampling, imputation, or summary aggregation may be useful engineering choices, but each makes choices about temporal resolution, which observations count, and how absences are represented.  A bedside monitor, a laboratory order, and a static admission field need not convey the same temporal or clinical information.  Moreover, the care process can affect both the values that are available and the chance that they are recorded.  A representation that ignores timing or observation patterns may discard clinically relevant structure; a representation that exploits them can improve prediction while also encoding care-process signals.  This tension makes irregular ICU data a methodological problem spanning statistical representation, workflow design, and interpretation rather than solely a model-selection problem.

For ICU analysis, missingness and irregularity matter at every layer of a study.  They influence which variables a representation can retain, which patterns a prediction model can exploit, which records satisfy a cohort definition, and whether a rule can assign an analytical label.  They also complicate causal adjustment: an absent measurement can represent no measurement, a measurement outside an available window, a source-data limitation, or a clinically informative decision not to measure.  Handling masks, time gaps, or sparse events can improve a predictive representation, but it does not by itself remove the interpretive consequences of the observation process.  The thesis consequently treats irregularity and informative measurement as motivations for careful representation and explicit limitations, rather than as local causal findings.

Two complementary needs follow.  First, an analysis must accommodate irregular longitudinal measurements rather than force all observations into an unqualified regular grid.  Architectures for sparse clinical series and missingness-aware recurrent models illustrate different ways to use values, time intervals, and observation patterns \cite{tipirneni2022strats,che2018grud}.  Second, the analysis needs interpretable variables that can connect a rich event stream to later questions about prognosis and adjusted mortality contrasts.  In this thesis, that intermediate layer consists of \emph{clinically inspired proxy states}: rule-derived analytical constructs that summarize evidence for selected clinical patterns from available measurements.

This wording is deliberate.  A rule-derived proxy label is not a chart-adjudicated diagnosis, ground truth, or a validated phenotype.  A positive tag records that the project-specific source rules were satisfied for the available data; a negative tag does not establish absence of a clinical condition.  Nevertheless, a transparent proxy-state layer can make the passage from irregular measurements to a structured downstream analysis inspectable.  It supplies common targets for multi-label prediction and a documented set of proxy-state exposures for observational analyses, while retaining a visible link to the measurements and rules from which each construct was derived.  Rule-based electronic phenotyping and weak-supervision research provide useful context for this design choice, but do not validate the local proxy definitions or their clinical meaning \cite{banda_2018_electronic_phenotyping,ratner_et_al_2020_snorkel}.

Prediction and causal analysis then answer different questions.  Predicting a rule-derived proxy label asks whether an irregular-time-series model can reproduce that label from the available inputs.  Estimating an adjusted mortality contrast for a proxy-state exposure asks a separate observational question about modeled differences in outcome under explicit assumptions.  Strong performance on the first task does not establish the clinical validity of the label, causal validity of the second analysis, or actionability of the exposure.  The latter task requires a stated causal model, project-specified directed acyclic graphs (DAGs), exposure-specific adjustment logic, attention to support and overlap, sensitivity and permutation diagnostics, and traceable evidence.  Even with these elements, adjusted estimates remain conditional on the adequacy of the proxy measurement, graph, temporal ordering, observed covariates, and model specification \cite{pearl_1995_causal_diagrams,hernan_robins_2016_target_trial,hernan_taubman_2008_well_defined_interventions}.

The framework is exercised on PhysioNet 2012 and MIMIC-III, two public ICU time-series resources that provide complementary settings for this problem \cite{silva2012physionet,johnson2016mimiciii}.  Both contain longitudinal ICU measurements and in-hospital mortality information, yet they differ in era, scale, available variables, data contracts, measurement processes, proxy-state definitions, and project graphs.  These differences make cross-dataset evaluation useful for examining the portability of an analytical workflow, but they preclude treating similarly named constructs as automatically equivalent or pooling numerical estimates without further justification.  The thesis therefore uses the datasets to test a shared methodological sequence while preserving dataset-specific definitions and interpretation boundaries.

This problem setting also explains why the thesis moves from data contracts to proxy states, prediction, and only then to adjusted analysis.  The same apparent clinical pattern may have different observable correlates when variable availability, sampling frequency, or coding conventions change.  If a downstream analysis hides those transformations, a reader cannot determine whether a difference arose from the source data, the rule-derived label, the prediction export, or the effect-estimation model.  Making these handoffs explicit is an engineering discipline with scientific consequences: it permits the analysis to be audited, identifies where limitations enter, and prevents an apparently seamless output from being interpreted as a seamless clinical fact.

\section{Thesis Gap and Objective}
\label{sec:introduction:thesis-gap-objective}

Prior work offers important components for this setting.  Irregular-time-series research provides representations and predictive methods for sparse, asynchronously observed medical data.  Electronic phenotyping provides strategies for constructing analytical variables when direct expert adjudication is unavailable \cite{banda_2018_electronic_phenotyping}.  Data programming formalizes programmatic heuristic labeling \cite{ratner_et_al_2016_data_programming}, while Snorkel provides related weak-supervision and downstream discriminative-training context \cite{ratner_et_al_2020_snorkel}.  Evaluations of medical LLM capabilities provide a bounded precedent for design-time clinical-knowledge elicitation \cite{singhal_et_al_2023_llm_clinical_knowledge}; work on LLM-derived causal-graph priors separately motivates their use as proposals rather than validated graphs \cite{darvariu_et_al_2024_llm_causal_graph_priors}.  Causal-inference research supplies DAG-based reasoning, double machine learning, heterogeneous-effect estimation, and diagnostics for observational studies \cite{chernozhukov2018dml,wager2018causalforest,athey2019grf}.  ICU-focused reviews further emphasize the difficulty of defining an actionable question, managing confounding, and distinguishing predictive patterns from treatment-effect claims in observational clinical records \cite{smit_2023_causal_inference_icu_scoping_review,bica_2021_individualized_treatment_effects_ehr_ml}.

The gap addressed here is an integration gap, not an assertion that any of these components is absent from the literature.  In practice, design-time knowledge elicitation, irregular-time-series prediction, proxy-state construction, and causal-effect estimation are often developed or evaluated separately.  Connecting them requires explicit interfaces: LLM-generated proposals must be distinguished from selected source-encoded designs; source-specific preprocessing must yield compatible records; rule-derived labels must be distinguishable from model predictions and algorithmic aggregates; prediction exports must be normalized before downstream use; and causal analyses must state how their exposures, covariates, graphs, diagnostics, and populations relate.  Without these interfaces, a model can appear technically successful while its design assumptions, labels, analysis rows, adjustment variables, or interpretation change silently between stages.

The objective of this thesis is therefore \emph{to construct and evaluate an evidence-tracked framework that combines LLM-assisted clinical and causal knowledge elicitation, deterministic proxy-label construction, deep prediction of proxy states from irregular ICU time series, and DAG-guided causal-effect estimation of in-hospital mortality}.  ``Evidence-tracked'' here means that the thesis distinguishes LLM design proposals, selected source-coded behavior, checked numerical artifacts, and unresolved provenance or validation requirements.  It is a methodological and engineering contribution accompanied by an empirical evaluation; it is not a claim of clinical validation, deployment readiness, or full clean-checkout computational reproducibility.

The framework has two distinct layers.  At design time, a structured LLM-assisted protocol elicited proposed proxy-state ontologies, decision-rule families, missingness considerations, and dataset-specific DAGs; project-selected designs were then encoded in deterministic tagger and graph source files.  During analytical execution, the sequence is dataset-specific preprocessing; deterministic generation of rule-derived proxy labels; multi-label proxy-state prediction; prediction normalization and algorithmic aggregation; exposure-specific adjustment under the project-specified DAGs; matching and heterogeneous-effect estimation; and robustness and provenance evaluation.  The LLM is not invoked in this executed patient-level pipeline.  The proxy states are not validated clinical diagnoses; the DAGs are not learned from data; no unconditional causal effect is claimed; and many illness-state proxies do not define well-specified interventions.  The workflow does not provide clinical recommendations, bedside decision support, or prospective evidence of benefit.

An integrated workflow is useful precisely because its interfaces can be inspected rather than assumed.  Dataset-specific preprocessing preserves the distinction between source records and analysis-ready objects.  Rule-based construction makes the origin of each proxy label visible.  Prediction normalization distinguishes probabilities from thresholded labels, and algorithmic aggregation distinguishes a deterministic vote from clinical consensus.  Project-specified DAGs and exposure-specific adjustment sets make the causal assumptions operationally visible, while matching and model-estimated CATE summaries provide different forms of observational evidence.  Finally, robustness and provenance checks separate a checked numerical statement from stronger claims about causal identification, clinical meaning, or rerun reproducibility.  The framework's purpose is not to erase these distinctions, but to retain them throughout the analysis.

This framing has practical consequences for how the results should be read.  A checked prediction metric concerns the archived evaluation of a model against rule-derived labels, not an independently verified clinical endpoint.  A matching output is a descriptive matched-pair outcome difference, not automatically a treatment effect.  A mean model-estimated CATE is an aggregate of fitted conditional estimates under the project assumptions, not an unconditional population quantity.  Sensitivity, permutation, and support diagnostics contribute evidence about particular failure modes, but none can substitute for a justified intervention definition, an adequate causal graph, or clinical review.  By placing these distinctions in the framework itself, the thesis avoids treating provenance, statistical modeling, and clinical interpretation as afterthoughts to an otherwise complete pipeline.

LLM-assisted clinical and causal knowledge elicitation is therefore a substantive design-time stage of CliniCause.  The structured protocol produced proposals for the proxy-state ontology, binary decision rules, missingness reasoning, and dataset-specific DAGs; the project selected designs and encoded them in active deterministic source code.  This role does not make the LLM a validated medical expert, establish that every proposal was implemented unchanged, or validate the proxy constructs or graphs.  Nor was the LLM an executed patient-level estimator: it did not preprocess records, generate runtime predictions, select adjustment variables dynamically, or estimate causal effects.  The distinction between design provenance and executable authority makes this contribution inspectable without treating generated proposals as clinical or causal evidence.

\section{Research Questions and Contributions}
\label{sec:introduction:research-questions-contributions}

The main research question is:

\begin{quote}
How can LLM-assisted clinical and causal knowledge elicitation, deterministic proxy-label construction, and deep modeling of irregular ICU time series be integrated into an evidence-tracked framework for DAG-guided adjusted effect estimation of in-hospital mortality?
\end{quote}

The word ``support'' is central.  The thesis investigates how a linked workflow can organize and qualify retrospective evidence; it does not claim to prove that proxy-state exposures cause mortality or that manipulating a proxy would improve outcomes.  The main question is addressed through seven secondary research questions:

\begin{description}
  \item[SRQ-1:] Which data contracts are needed to connect PhysioNet 2012, MIMIC-III, the irregular-time-series prediction workflow, and the causal-analysis pipeline?
  \item[SRQ-2:] How were LLM-assisted proposals for dataset-specific proxy ontologies, decision rules, missingness considerations, and DAGs selected, encoded, and traced to active source?
  \item[SRQ-3:] How do STraTS and the included sequence-model baselines perform on multi-label proxy-state prediction across MIMIC-III and PhysioNet?
  \item[SRQ-4:] How are predicted proxy states normalized and aggregated before downstream analysis?
  \item[SRQ-5:] What observed adjustment sets are selected by the dataset-specific, project-specified DAGs for proxy-state exposure analyses?
  \item[SRQ-6:] What matching, model-estimated CATE, sensitivity, and permutation evidence is available for the prespecified proxy-state exposures?
  \item[SRQ-7:] What limitations follow from proxy labels, intervention definition, overlap, measurement error, model dependence, provenance, and unmeasured confounding?
\end{description}

The primary contribution is an end-to-end, evidence-tracked framework linking LLM-assisted clinical and causal knowledge elicitation, irregular ICU preprocessing, deterministic proxy-label construction, deep proxy-state prediction, algorithmic aggregation, DAG-guided adjustment, matching, heterogeneous-effect estimation, and robustness diagnostics.  Its value is the explicit connection between these layers: prompt and source artifacts expose how a proposed construct becomes an implemented rule, while a shared proxy-state interface makes it possible to inspect how that construct moves from rule derivation to prediction and then to an observational exposure analysis.

Secondary methodological contributions are a traceable LLM-guided design layer for dataset-specific proxy definitions, decision rules, and DAG proposals; explicit dataset-specific data contracts between the predictive and causal repositories; source-coded, exposure-specific DAG adjustment logic; and a multi-estimator analysis comprising CausalForestDML, LinearDML, and exploratory CausalPFN.  The framework also combines matching, sensitivity, permutation, and support diagnostics so that estimator outputs are interpreted alongside their available empirical checks.  CausalForestDML is the primary model-estimated causal-analysis estimator, LinearDML is a secondary comparator, and CausalPFN is retained as an exploratory complementary estimator because it lacks the complete diagnostic support available for the DML estimators.

The empirical contribution is a checked comparison of five predictive model families across the two datasets, original-cohort adjusted-effect analyses for every prespecified dataset-specific proxy-state exposure, cross-estimator directional comparison, separate outcome-downsampling robustness analyses, and cross-dataset evaluation without numerical pooling.  The thesis also contributes reproducibility and evidence infrastructure: checked numerical tables, a results manifest, checksums, a source packet, a decision register, corrected result figures, and explicit tracking of unresolved provenance.  These materials support numerical traceability, but they do not establish that the analysis can be rerun from a clean checkout or that every historical producing setting has been recovered.

Together, these contributions are intended to be cumulative rather than interchangeable.  An LLM-generated proposal alone would not establish an implemented or clinically valid design.  A predictive comparison alone would not explain how its targets become downstream analytical variables.  A causal estimator alone would not document how irregular measurements, proxy definitions, or graph-selected covariates entered the analysis.  Conversely, an auditable interface does not turn a rule-derived label into a clinical diagnosis or a graph-guided contrast into a well-defined intervention effect.  The contribution hierarchy therefore places the integrated framework first, its LLM-assisted design and shared proxy/data-contract layers next, empirical comparisons and estimator triangulation after them, and evidence-tracking infrastructure alongside every layer.  This ordering keeps the thesis focused on a transparent methodological connection while leaving clinical validation and stronger causal questions as necessary future work.

\begin{table}[htbp]
\centering
\caption{Contribution hierarchy, evidence supplied by the thesis, and principal boundaries.}
\label{tab:introduction-contributions}
\small
\begin{tabular}{p{0.25\textwidth}p{0.40\textwidth}p{0.25\textwidth}}
\toprule
Contribution & Evidence supplied by the thesis & Principal boundary \\
\midrule
End-to-end methodological framework & Linked preprocessing, proxy, prediction, aggregation, graph, estimation, and diagnostic chapters. & Integration is demonstrated as a retrospective research workflow, not clinical validation. \\
LLM-guided design layer & Preserved prompts and run exports linked at schema level to deterministic tagger and graph sources. & Model settings, human selection, line-level traceability, clinical review, and graph correctness remain incomplete. \\
Shared proxy-state interface & Source-defined rule labels, prediction exports, aggregation, and proxy-state exposure analyses. & Constructs are rule-derived proxies, not chart-adjudicated diagnoses or ground truth. \\
Cross-dataset predictive evaluation & Checked comparison of five model families on MIMIC-III and PhysioNet. & Rankings are dataset specific; target validity and export lineage remain qualified. \\
DAG-guided multi-estimator analysis & Dataset-specific graphs, adjustment logic, matching, CausalForestDML, LinearDML, and exploratory CausalPFN. & Graph correctness, intervention definition, overlap, and unmeasured confounding are not established. \\
Robustness and diagnostic framework & Matching, sensitivity, permutation, and outcome-downsampling analyses. & Diagnostics qualify results; they do not confirm causal identification. \\
Evidence tracking and reproducibility infrastructure & Checked tables, manifest, checksums, source packet, decision register, and corrected figures. & Numerical traceability is stronger than clean-checkout computational reproducibility. \\
\bottomrule
\end{tabular}
\end{table}
\clearpage

The completed empirical record provides a concise orientation for the chapters that follow.  The checked predictive comparison included STraTS, GRU, GRU-D, TCN, and SAnD.  STraTS led the four archived test metrics in MIMIC-III, whereas GRU-D led them in PhysioNet; this dataset-dependent pattern is not a claim of universal model superiority.  In original-cohort adjusted analyses, CausalForestDML and LinearDML agreed in effect direction for all 19 dataset--exposure comparisons, while all three estimators agreed in direction for 18 of 19 comparisons.  The exploratory CausalPFN analysis reproduced the prevailing direction in nearly every original-cohort comparison, indicating promise as a complementary estimator while lacking the complete diagnostic support available for the DML estimators.  Matching and robustness diagnostics provide qualified supporting evidence, and outcome-downsampled analyses are treated as a separate robustness population rather than pooled with the original cohorts.

These findings are deliberately directional and high level.  They do not provide exposure-level effect claims, significance conclusions, or treatment priorities.  They instead motivate a careful reading of the later methods, results, and discussion: agreement across fitted estimators is useful triangulation, but it does not validate the proxy states, establish that the project DAGs are correct, eliminate unmeasured confounding, establish positivity, or demonstrate bedside utility.  The appropriate interpretation is an evidence-bounded account of what an integrated retrospective workflow can contribute and what further clinical, causal, and provenance work remains necessary.

\section{Thesis Organization}
\label{sec:introduction:thesis-organization}

The remainder of the thesis follows the analytical chain from context to interpretation.  Chapter 2 reviews irregular medical time-series modeling, LLM-assisted knowledge elicitation, clinical proxy construction, weak supervision, and observational causal-inference concepts that frame the study.  Chapter 3 separates the design-time knowledge-construction layer from the executed analytical layer and formalizes the study units, prediction and causal questions, proxy-state exposures, outcome, and assumptions.  Chapters 4--6 build the data, proxy, and prediction layers: Chapter 4 describes dataset-specific preprocessing and data contracts, Chapter 5 documents LLM-assisted clinical proxy-state and decision-rule construction, and Chapter 6 presents predictive modeling of rule-derived proxy labels.

Chapters 7--9 define how the downstream observational analyses are governed.  Chapter 7 describes the project-specified causal graphs, adjustment logic, matching, and heterogeneous-effect estimators.  Chapter 8 specifies robustness, sensitivity, permutation, and validation design.  Chapter 9 records the experimental matrix, original and outcome-downsampled populations, and reproducibility boundary.  Chapter 10 then reports the checked numerical results using the approved estimator and population hierarchy.  Chapter 11 interprets the research-question answers, cross-dataset patterns, and limitations.  Finally, Chapter 12 summarizes the thesis contributions and identifies future work needed to strengthen clinical validity, causal interpretation, reproducibility, and translation.
