\chapter{Problem Definition and Study Design}
\label{chap:problem-definition-study-design}

\section{Units, Time Horizons, and Data Objects}
\label{sec:problem-definition-study-design:units-time-horizons-data-objects}
% Section ID: C3.1.
% Evidence map: see thesis-writing/planning/chapter_evidence_map.md.

This thesis studies retrospective intensive-care records represented as irregular clinical time series.  The operational study unit is a time-series record or ICU stay, depending on the source dataset and preprocessing path.  PhysioNet 2012 source files use record identifiers, whereas MIMIC-III source tables use ICU-stay and admission identifiers.  The implemented pipelines normalize these source-specific identifiers to the canonical join key \texttt{ts\_id}.  This identifier should be read as the analysis-record identifier used by the software; it is not assumed to be a globally unique person identifier across repeated admissions or stays.

For a study unit \(i\), the observed longitudinal data are represented as a finite collection of measurement events
\[
  \mathcal{O}_i = \{(i,t_{ij},v_{ij},x_{ij}) : j=1,\ldots,n_i\},
\]
where \(i\) is the normalized \texttt{ts\_id}, \(t_{ij}\) is elapsed time in minutes from record start or ICU admission, \(v_{ij}\in\mathcal{V}\) is the observed variable name, and \(x_{ij}\) is the numeric value recorded for that variable.  In the causal repository this long-format object is stored in \texttt{ts} with the required columns \texttt{ts\_id}, \texttt{minute}, \texttt{variable}, and \texttt{value}.  Static or baseline quantities such as age, sex or gender coding, height, weight, and ICU-type indicators are represented within the same event schema as variables observed at or near the beginning of the record when the preprocessing implementation provides them.

The event representation is intentionally irregular.  Observations occur at nonuniform times, different variables are measured with different frequencies, and most variables are absent at most possible times.  A missing measurement is therefore not interpreted as a normal value.  It may reflect a clinical decision not to measure, an unrecorded observation, a source-table limitation, or a preprocessing exclusion.  The thesis treats informative measurement and missingness processes as important threats to interpretation, not as causal findings established by the repository.

The primary patient- or stay-level outcome used by the downstream causal repository is \texttt{in\_hospital\_mortality}.  In the causal processed artifact, this outcome is stored in \texttt{oc}, the outcome table keyed by \texttt{ts\_id}.  The third object in that processed artifact, \texttt{ts\_ids}, is the sorted collection of normalized identifiers represented in \texttt{ts}.  Later predictive processing uses a distinct split-aware artifact and must not be conflated with this causal contract.

\section{Design-Time Knowledge Construction and Executed Analysis}
\label{sec:problem-definition-study-design:design-time-executed-layers}

The framework distinguishes a design-time knowledge-construction layer from the patient-level analytical layer.  At design time, the official dataset-description link and instructions to research relevant clinical, causal, and time-series literature were supplied to a structured LLM-assisted elicitation protocol.  The protocol requested a dataset audit, proxy-state ontology, binary decision-rule families, missingness and measurement-process reasoning, and a dataset-specific DAG.  The resulting proposals were project-selected and encoded in deterministic tagger and graph source files.  Available evidence does not establish that every proposal was implemented unchanged, a complete line-by-line prompt-output-to-code mapping, or a formal clinician-review chain.

The analytical layer begins only after those designs have been encoded.  It preprocesses raw ICU data, applies deterministic decision-tree rules to generate rule-derived proxy labels, trains deep irregular-time-series models to predict those labels, normalizes predicted outputs, and forms the archived final aggregate from one rule-derived source and four predicted-label sources before DAG-guided matching, effect estimation, sensitivity analysis, and permutation diagnostics.  No LLM is invoked for patient-level preprocessing, label generation, prediction, adjustment selection, or effect estimation during this executed layer.

\begin{table}[htbp]
\centering
\caption{Separation of design-time knowledge construction from executed patient-level analysis.}
\label{tab:study-design-layers}
\small
\begin{tabular}{@{}p{0.18\textwidth}p{0.48\textwidth}p{0.24\textwidth}@{}}
\toprule
Layer & Sequence & Evidence boundary \\
\midrule
Design time & Official dataset-description link and instructions to research relevant clinical, causal, and time-series literature $\rightarrow$ structured LLM elicitation $\rightarrow$ proposed ontology, rules, missingness considerations, and DAG $\rightarrow$ project selection and source-code encoding. & Prompts document proposals; active source defines implementation; formal clinical review remains incomplete. \\
Executed analysis & Raw data $\rightarrow$ preprocessing $\rightarrow$ deterministic rule-derived labels $\rightarrow$ deep prediction $\rightarrow$ prediction normalization and mixed-source aggregation $\rightarrow$ DAG-guided estimation and diagnostics. & No runtime LLM call; every downstream inference remains conditional on proxy, graph, support, and model assumptions. \\
\bottomrule
\end{tabular}
\end{table}

\section{Prediction Task}
\label{sec:problem-definition-study-design:prediction-task}
% Section ID: C3.2.
% Evidence map: see thesis-writing/planning/chapter_evidence_map.md.

The prediction task is a multi-label proxy-state prediction problem over irregular ICU time series.  For each study unit \(i\) and proxy state \(k\), let \(L^{rule}_{ik}\in\{0,1\}\) denote a rule-derived proxy label.  These labels are clinically inspired analytical constructs generated from observed data by the deterministic source-code rules selected and encoded after LLM-assisted design elicitation.  They are weak clinical labels or proxy phenotypes, not chart-adjudicated diagnoses, manually verified clinical states, or validated phenotypes unless separate validation evidence is supplied in a later chapter.

The supervised predictive workflow estimates a vector of proxy-state labels from the observed time-series representation.  Its target matrix is loaded from a latent-label CSV keyed by \texttt{ts\_id}; the supervised proxy-label targets are not taken from the mortality column in \texttt{oc}.  For model \(m\), the exported prediction object may contain probabilities \(\hat p^{(m)}_{ik}\) and corresponding binary predicted labels \(\hat L^{(m)}_{ik}\).  These outputs are model estimates of rule-derived proxy labels rather than direct measurements of patient physiology.

Within the executed layer, five analytical tasks are distinguished.  First, deterministic proxy-state construction derives binary rule-derived proxy labels from observed clinical measurements.  Second, multi-label proxy-state prediction estimates those rule-derived labels from irregular time-series inputs.  Third, the archived final causal runs aggregate one rule-derived table with four aligned binary model-prediction tables into a majority-vote proxy-label table.  Fourth, mortality prediction or association analysis evaluates whether proxy-state representations contain prognostic information about \texttt{in\_hospital\_mortality}.  Fifth, observational causal-effect estimation treats selected columns from that mixed rule-and-prediction aggregate as modeled exposures and estimates adjusted contrasts under explicit assumptions.  Detailed proxy rules, model architectures, estimator algorithms, and numerical results are intentionally deferred to later chapters.

\section{Causal Question, Exposures, Outcome, Estimands, Assumptions}
\label{sec:problem-definition-study-design:causal-question-exposures-outcome-estimands-assumptions}
% Section ID: C3.3.
% Evidence map: see thesis-writing/planning/chapter_evidence_map.md.

The downstream causal-analysis setting is retrospective and observational.  No intervention is assigned by this study.  The repository uses software parameters named \texttt{treatment} for selected binary \texttt{LAT\_*} columns, but many such columns represent illness-state or organ-dysfunction proxy states rather than well-defined administered therapies.  In the thesis narrative these variables are therefore described as proxy-state exposures or modeled treatment variables unless a later section explicitly defines a manipulable intervention.

Let \(A_i\in\{0,1\}\) denote a selected proxy-state exposure, \(Y_i\) denote the binary outcome \texttt{in\_hospital\_mortality}, \(W_i\) denote observed adjustment variables selected from a project-specified DAG and available columns, and \(X_i\) denote effect-modifier features used to describe heterogeneity.  The dataset-specific DAGs were developed through a project design process that included LLM-assisted clinical and causal elicitation and were then encoded in source code; the implemented graph scripts, not the prompt outputs alone, define the DAG artifacts used by the pipeline.  Potential-outcome notation such as \(Y_i(a)\) is useful as conceptual language, but causal interpretation requires more than the presence of a graph or an estimator.  It depends on assumptions about the causal graph, consistency, exchangeability conditional on observed covariates, positivity or overlap, outcome and exposure measurement, and unmeasured confounding \cite{pearl_1995_causal_diagrams,hernan_robins_2016_target_trial,hernan_taubman_2008_well_defined_interventions}.

Accordingly, later estimates should be interpreted as adjusted effect estimates under project assumptions, not as unconditional clinical effects.  The existence of a DAG, matching script, or CATE estimator does not prove identification.  Proxy states may be useful for structured analysis while still carrying label noise, temporal ambiguity, and intervention-definition limitations.

The main research question is how LLM-assisted clinical and causal knowledge elicitation, deterministic proxy-label construction, and deep modeling of irregular ICU time series can be integrated into an evidence-tracked framework for DAG-guided adjusted effect estimation of in-hospital mortality.  The seven secondary questions stated in Chapter~\ref{chap:introduction} cover data contracts; selection, encoding, and traceability of LLM-assisted proxy and DAG proposals; five-family predictive performance; prediction normalization and aggregation; DAG-selected adjustment sets; matching and model-estimated CATE evidence with available diagnostics; and the limitations that bound those findings.  These are study questions and design commitments; their answers are reported in Chapters~\ref{chap:results} and~\ref{chap:discussion}.
