\chapter{Predictive Modeling of Proxy States}
\label{chap:predictive-modeling-proxy-states}

This chapter describes the predictive layer that estimates rule-derived proxy-state vectors from ICU time-series records.  Its supervision chain is LLM-assisted decision-rule design, project selection and deterministic source encoding, generation of rule-derived proxy labels, and deep-model prediction of those labels.  The supervised target is the resulting rule-derived proxy-label matrix loaded from a separate CSV file keyed by \texttt{ts\_id}; the inputs are irregular measurement histories plus static covariates prepared under the data contracts described in Chapter~\ref{chap:data-preprocessing}.  The models therefore predict analytical proxy labels rather than adjudicated clinical states.

The final comparison contains four learned models: STraTS, GRU, GRU-D, and TCN.  This chapter defines their representation, training, evaluation, and export mechanics; experiment selection and numerical comparisons are deferred to Chapters~\ref{chap:experimental-design} and~\ref{chap:results}.

\section{Self-Supervised STraTS Pretraining}
\label{sec:predictive-modeling-proxy-states:self-supervised-strats-pretraining}
% Section ID: C6.1.
% Evidence map: see thesis-writing/planning/chapter_evidence_map.md.

STraTS represents an irregular time series as a set of observed measurement tokens rather than as a fully imputed grid.  The implemented STraTS class constructs each token by adding three learned components: a continuous embedding of measurement time, a continuous embedding of the normalized measurement value, and an embedding of the variable identity.  The token sequence is contextualized with a transformer-style block and then pooled with fusion attention to obtain a fixed-dimensional time-series representation \cite{tipirneni2022strats}.  Static covariates are combined with this representation before either the self-supervised forecasting head or the supervised proxy-label head is applied.

In the verified STraTS implementation, contextualized observation embeddings are pooled and combined with a learned embedding of static covariates.  The recurrent, convolutional, and decay baselines in the final comparison do not have a supported self-supervised pretraining path in the current source.

The pretraining dataset loader reads the split-aware processed artifact and removes held-out test identifiers before constructing unsupervised examples.  It builds the variable vocabulary from variables observed in the pretraining train split, derives value means and standard deviations from that train split, and saves the resulting variable list, normalization table, and maximum time horizon to \texttt{pt\_saved\_variables.pkl}.  For PhysioNet 2012 the maximum time is 48 hours.  For MIMIC-III the loader permits observations up to five days but samples at most a 24-hour input context and clamps implausibly old age values in the same way as the supervised loader.

Each self-supervised example is formed by sampling a forecast anchor from observed timestamps that are not the final timestamp and are at least 12 hours after record start.  Observations up to the anchor form the historical context, truncated by the maximum observation count and, for MIMIC-III, by the 24-hour context limit.  The target window extends two hours after the anchor.  For each variable observed in that future window, the implementation keeps the last observed normalized value and sets the corresponding forecast mask entry to one; variables without future observations are masked out of the loss.  Input times are rescaled to the interval \([-1,1]\) using the dataset-specific maximum time.

The pretraining head maps the combined time-series and static representation to one forecast value per variable.  Its objective is masked squared error over variables observed in the sampled future window, so unobserved future variables do not contribute to the loss.  The pretraining evaluator materializes batches for each evaluation split, sampling three forecast batches per evaluation chunk when a split is first evaluated, and reports \texttt{loss\_neg}, the negative masked loss, for checkpoint selection.  The best pretraining checkpoint is written as \texttt{checkpoint\_best.bin}; loss values themselves are not reported in this methods chapter.

Checkpoint-loaded supervised training is implemented by constructing the target supervised architecture, loading the supplied pretraining checkpoint, and copying overlapping tensors into the current model state before calling \texttt{load\_state\_dict}.  When \texttt{--load\_ckpt\_path} is supplied, the supervised dataset also loads \texttt{pt\_saved\_variables.pkl} from the same directory to reuse the pretraining variable vocabulary, normalization statistics, and maximum-time metadata.  This source path supports warm-started STraTS-family training, but the archived final exports still lack a complete manifest linking each exported CSV to the intended pretraining checkpoint, supervised checkpoint, split artifact, and archive-copy step.

\section{Supervised Multi-Label Models}
\label{sec:predictive-modeling-proxy-states:supervised-multi-label-models}
% Section ID: C6.2.
% Evidence map: see thesis-writing/planning/chapter_evidence_map.md.

The supervised prediction task uses a deterministic rule-derived latent-label CSV as its target matrix.  The loader accepts identifier columns named \texttt{ts\_id}, \texttt{icustay\_id}, or \texttt{ICUSTAY\_ID}, canonicalizes them to \texttt{ts\_id}, and checks consistency when more than one accepted identifier column is present.  It then filters the label table to the supervised split identifiers, raises an error for missing labels, raises an error for duplicate normalized identifiers, sorts rows to match the internal supervised identifier order, and treats every non-\texttt{ts\_id} column as a proxy-label target.  The number and order of supervised targets are therefore determined at runtime by the CSV schema, but their semantic status remains rule-derived rather than clinical ground truth.

All supervised backends share a multi-label output contract.  During training the model returns binary cross-entropy with logits against the full target vector, using per-target positive-class weights computed from the supervised train split.  During evaluation or prediction export, the same model call without labels returns sigmoid probabilities.  Static covariates are normalized from training rows and combined with each model's time-series representation.  In scratch training the binary head maps the combined representation directly to proxy-label logits; checkpoint-loaded STraTS training uses an intermediate forecasting head so overlapping pretraining tensors can be reused.

The STraTS supervised backend uses the irregular observation triplets described in Section~\ref{sec:predictive-modeling-proxy-states:self-supervised-strats-pretraining}: time, value, and variable embeddings, transformer contextualization, fusion attention, and static-feature combination \cite{tipirneni2022strats}.

The GRU baseline uses a dense hourly grid rather than the raw irregular token set.  For each hour and variable, the loader builds value, observation-mask, and elapsed-time-since-observation channels, mean-fills unobserved values from training data, normalizes the filled values, and concatenates the three channel groups before passing them to a gated recurrent unit \cite{cho2014gru}.  The final recurrent hidden state is concatenated with the static-feature embedding and projected through the shared supervised head.

The GRU-D backend uses a model-specific irregular sequence representation containing observed values, missingness masks, elapsed-time tensors, sequence lengths, and static covariates \cite{che2018grud}.  The implemented cell updates last observed values, applies learned exponential decay to inputs and hidden states as a function of elapsed time, and includes missingness masks in the recurrent gates.  The representation passed to the supervised head is the final valid recurrent state for each sequence combined with the static-feature embedding.

The TCN baseline uses the same dense hourly value, observation-mask, and elapsed-time representation as the GRU baseline, but processes the sequence with temporal convolutional residual blocks \cite{bai2018tcn}.  The implementation permutes the dense grid into channel-first form, applies a temporal convolutional network, takes the last temporal embedding, and concatenates it with the static-feature embedding before projection to proxy-label logits.

Together, these backends compare sparse-event attention, ordinary recurrence on a prepared grid, missingness-aware recurrence with learned decay, and temporal convolution.  All four have archived prediction outputs for both datasets, while checkpoint-to-export, split, and archive-copy lineage remains incomplete.  Table~\ref{tab:background-model-families} summarizes their inductive biases without ranking them by sophistication.

The supervised evaluator computes metrics independently for each rule-derived target column in the requested split.  Target columns with only one observed class in that split are skipped.  For the remaining targets, it computes AUROC, area under the precision-recall curve, and the maximum value of the pointwise minimum of precision and recall.  The reported split-level values are simple averages over the non-degenerate target columns, with loss and negative loss also recorded.  During supervised training, checkpoint selection uses the validation value of \texttt{auprc + auroc}; this validation score should not be interpreted as a test score.  These metrics quantify learnability or reproducibility of rule-derived targets, not clinical construct validity or correctness of the LLM-assisted decision-rule design.  A model can reproduce both useful rule structure and systematic rule error.

Training mechanics are shared across the supervised backends at the top-level script.  The script constructs a deterministic seed by adding the run index to the configured seed, uses AdamW for trainable parameters, clips gradient norms at 0.3, evaluates at the configured validation schedule, and saves \texttt{checkpoint\_best.bin} when the validation selection metric improves.  If an \texttt{eval\_train} split is evaluated, it is the first 2000 training examples in the current supervised split object, which makes it a diagnostic subset rather than a separate external evaluation set.

\section{Prediction Export and Normalization}
\label{sec:predictive-modeling-proxy-states:prediction-export-normalization}
% Section ID: C6.3.
% Evidence map: see thesis-writing/planning/chapter_evidence_map.md.

Prediction export is requested with \texttt{--save\_pred\_csv\_path}.  The export routine can run after a training run or in an export-only invocation with no training steps scheduled.  Immediately before export, the script reloads \texttt{checkpoint\_best.bin} from the current output directory only if that file exists.  The exported split is selected by \texttt{--predict\_split}, whose accepted values are \texttt{train}, \texttt{val}, \texttt{test}, and \texttt{all}; in the \texttt{all} case the exporter iterates over all supervised identifiers in the train, validation, and test split object.  The inspected wrapper scripts and archived export logs support \texttt{predict\_split=all} for the final prediction CSVs.  The logs also record \texttt{seed=2023}, and archived \texttt{checkpoint\_best.bin} files can be hashed; nevertheless, the records do not supply split identifiers, a processed-artifact hash, or an independently verified checkpoint-to-export and archive-copy manifest.

For every exported row, the exporter removes labels from the batch, obtains probability outputs from the model, thresholds each probability at 0.5, and writes a combined CSV.  The first column is \texttt{ts\_id}.  It is followed by one probability column named \texttt{<label>\_prob} for each target and then one thresholded binary prediction column named exactly as the target label.  These binary columns are model-predicted proxy labels, not rule-derived labels and not clinical adjudications.

The combined prediction CSV is not the same file shape required by downstream voter aggregation.  The normalization helper validates \texttt{ts\_id}, separates probability and binary fields, and writes binary-only voter files.  The majority-vote contract then requires a shared identifier and binary proxy schema in the approved order.  The archived final vote combines one rule-derived table with predicted-label tables from GRU, GRU-D, STraTS, and TCN; Chapter~\ref{chap:clinical-proxy-state-construction} defines its semantics.  This is an algorithmic handoff, not expert consensus.
