\chapter{Discussion}
\label{chap:discussion}

This chapter interprets the evidence reported in Chapter~\ref{chap:results} within the study design and assumptions established in Chapters~\ref{chap:problem-definition-study-design}--\ref{chap:experimental-design}.  The central distinction is between what the thesis evidence demonstrates empirically---an implemented, evidence-tracked workflow and stable patterns across several analyses---and what it does not establish clinically or causally.  The discussion therefore treats the reported quantities as prediction results, descriptive matched-pair comparisons, or adjusted model estimates under project assumptions, according to their source.

\section{Answering the Research Questions}
\label{sec:discussion:answer-research-questions}
% Section ID: C11.1.
% Evidence map: see thesis-writing/planning/chapter_evidence_map.md.

\subsection{Main Research Question}
\label{subsec:discussion:main-rq}

The main research question asked: \emph{How can LLM-assisted clinical and causal knowledge elicitation, deterministic proxy-label construction, and deep modeling of irregular ICU time series be integrated into an evidence-tracked framework for DAG-guided adjusted effect estimation of in-hospital mortality?}

The direct answer is that this thesis demonstrates an evidence-tracked methodological framework with a sequence of explicit interfaces.  At design time, a structured LLM-assisted protocol generated proposals for dataset-specific proxy ontologies, decision rules, missingness considerations, and DAGs; project-selected designs were encoded in deterministic source.  During execution, dataset-specific irregular measurements are converted into normalized data contracts; the source-coded rules create rule-derived proxy labels; irregular-time-series models predict those labels; and probabilities and binary predicted fields are exported and normalized.  In every archived final causal run, the vote then combines one rule-derived table with four model-prediction tables.  Source-coded, project-specified DAGs define exposure-specific adjustment logic, and matching and heterogeneous-effect estimators analyze the resulting mixed-source aggregate.  Robustness and provenance checks qualify the interpretation.  The framework was implemented and empirically exercised across MIMIC-III and PhysioNet 2012 without a runtime LLM call.

This is a feasibility and integration result, not a conclusive solution to the clinical causal question or an empirical evaluation of LLM quality.  The proxy states are not chart-adjudicated diagnoses, many exposures describe illness states rather than well-defined interventions, and the LLM-assisted project DAGs are assumed rather than empirically established.  The adjusted estimates remain conditional on construct measurement, consistency, exchangeability, temporal ordering, overlap, and model adequacy.  The framework therefore supports auditable retrospective analysis and hypothesis generation, but it does not produce clinical treatment recommendations.  In summary, the thesis demonstrates the feasibility of linking LLM-assisted knowledge construction, irregular-time-series representation, proxy-state prediction, and assumption-guided adjusted effect estimation in one auditable workflow, while showing that the strength of causal interpretation is constrained chiefly by construct validity, intervention definition, causal-graph validity, empirical support, and provenance.

\subsection{Data Contracts and End-to-End Pipeline Integration}
\label{subsec:discussion:data-contracts-integration}

\paragraph{SRQ-1---data contracts.}
The pipeline requires at least two distinct processed-artifact contracts.  The causal repository uses \texttt{[ts, oc, ts\_ids]}, with a normalized \texttt{ts\_id} linking the long event table to the outcome table.  The STraTS and baseline workflow uses a split-aware artifact with events, outcomes, and explicit train, validation, and test identifier collections.  The router and conversion logic bridge these contracts by normalizing identifiers and constructing the split-aware form when requested.  Prediction exports add another interface: \texttt{ts\_id}, paired probability and binary proxy-state columns, and subsequently binary-only voter files.  The answer is therefore \emph{implementation supported}: the repository defines and validates these boundaries, but exact processed-pickle hashes, producing commands, and final split lineage remain incomplete.

\paragraph{SRQ-4---prediction normalization and aggregation.}
Prediction outputs are normalized into paired probability and binary proxy-state fields, while downstream voter inputs retain the binary fields in an approved order.  Majority voting aligns shared normalized identifiers, checks identical binary proxy-state schemas, and applies the implemented tie rule.  In every archived final run, the logs enumerate one deterministic rule-derived source and four thresholded prediction sources---GRU, GRU-D, STraTS, and TCN---with SAnD absent from the vote.  This supplies a deterministic mixed-source handoff into the causal repository.  It is algorithmic aggregation rather than clinical consensus; the logs retain external filenames but do not co-archive the voter-file bytes or hashes, while checkpoint mappings and archive-copy provenance remain incomplete.  Moreover, aggregation cannot automatically correct systematic error shared by models trained on the same rule-derived targets, especially when the originating rule source itself participates in the vote.

\subsection{Proxy-State Construction, Prediction, and Aggregation}
\label{subsec:discussion:proxy-prediction-aggregation}

\paragraph{SRQ-2---LLM-assisted design selection, encoding, and traceability.}
The preserved generic and dataset-specific prompts, final run exports, active tagger and graph files, and generated proxy and DAG artifacts establish a traceable design sequence at artifact and schema level.  The final prompt outputs align with the active eleven-state PhysioNet and ten-state MIMIC schemas, while deterministic source rules and stable \texttt{LAT\_*} columns make implemented behavior inspectable.  The answer remains \emph{partially traceable and unresolved clinically}: complete accepted/rejected decision records and line-by-line prompt-output-to-code correspondence are unavailable; the two datasets use different ontologies and measurements; threshold rules may misclassify constructs; clinical review remains incomplete; and no chart-adjudicated reference labels were archived.  These are LLM-assisted, project-selected proxy designs encoded as deterministic rules, not diagnoses, ground truth, or validated phenotypes.  The rule-to-label transition is conceptually related to programmatic labeling, while the aggregation layer does not implement a learned Snorkel label model \cite{ratner_et_al_2016_data_programming,ratner_et_al_2020_snorkel}.

\paragraph{SRQ-3---predictive modeling.}
The validated comparison included STraTS, GRU, GRU-D, TCN, and SAnD.  STraTS led the four archived test metrics in MIMIC-III, whereas GRU-D led them in PhysioNet.  The leading architecture therefore differed by dataset.  This is empirically supported evidence that representation choice is dataset dependent, not evidence for universal model superiority.  Handling irregularity or missingness can be useful in several ways, and prior work distinguishes sparse-token representations from missingness-aware recurrent processing \cite{sun_2026_review_irregular_medical_timeseries,lipton_kale_wetzel_2016_missingness_rnns,tipirneni2022strats,che2018grud}.  The local comparison does not reproduce the original papers' benchmark tasks exactly.

Several mechanisms may contribute to the dataset-specific pattern: variable availability, measurement frequency, missingness, label prevalence, proxy definitions, preprocessing, dataset era, sample size, and ICU composition.  The study did not isolate those mechanisms, so they remain possible explanations rather than findings.  In addition, no uncertainty intervals or paired between-model tests were archived, checkpoint-to-export provenance is incomplete, and the prediction targets themselves lack clinical validation.  Statistical superiority is therefore not established.

\subsection{Adjusted Effect Estimation and Estimator Agreement}
\label{subsec:discussion:adjusted-effects-agreement}

\paragraph{SRQ-5---DAG-guided adjustment.}
Dataset-specific, source-coded DAGs select exposure-specific observed adjustment sets from columns available in each analysis frame.  LLM-assisted elicitation was a substantive design-time input, while the active graph code is the implementation authority.  The graphs were neither learned from the data nor clinically or empirically validated, and the adjustment sets differ across exposures.  Several validated rows contain empty observed adjustment sets, while treatment-specific identifiability remains unresolved throughout the result sources.  The answer is thus \emph{conditional on assumptions}: a computed set documents how the project applied its graph, but it does not prove exchangeability or the absence of unmeasured confounding \cite{pearl_1995_causal_diagrams}.

\paragraph{SRQ-6---matching, CATE, sensitivity, and permutation evidence.}
The approved hierarchy uses original cohorts as primary, CausalForestDML as the primary estimator, LinearDML as secondary, CausalPFN as exploratory, and outcome-downsampled analyses as robustness evidence.  In the primary forest results, all nine MIMIC original-cohort mean model-estimated CATE summaries were positive; nine of ten PhysioNet summaries were positive, with shock negative.  This pattern supports the narrower conclusion that the project-defined proxy-state exposures contain systematic mortality-related variation after the implemented adjustment procedure, that patterns differ by construct and dataset, and that heterogeneous-effect estimators can be applied to the constructed analysis objects.  It does not show that a proxy state caused death, that changing it would change mortality, or that a ranking identifies intervention priorities.

CausalForestDML and LinearDML agreed in direction for all 19 original-cohort dataset--exposure pairs.  This complete directional concordance is useful model-form triangulation: the sign pattern was not unique to the nonlinear forest final stage.  It does not imply equal magnitudes, orderings, estimands, uncertainty, or correctness, and it does not validate the DAG or rule out residual confounding.  Indeed, Chapter~\ref{chap:results} shows that magnitudes and within-dataset orderings can differ even when signs agree.

Within this pipeline, CausalPFN showed promise as a complementary estimator because it reproduced the prevailing direction across nearly all prespecified comparisons.  It contributed to all-three directional concordance for 18 of 19 original-cohort dataset--exposure pairs despite using a different modeling approach.  This positive result is bounded: magnitudes were not identical to DML magnitudes, the archived sensitivity and permutation stages were intentionally absent, and the approved literature corpus lacks a primary method citation.  CausalPFN therefore remains an exploratory empirical contribution rather than an estimator with the same diagnostic support as the DML analyses.

PhysioNet shock is the central exception rather than a detail to be averaged away.  CausalForestDML and LinearDML were negative, CausalPFN was slightly positive, and the descriptive matching summary was positive.  This disagreement is a substantive caution about model, representation, and support sensitivity.  It should not be resolved by a vote among estimators, and the negative DML estimates should not be described as a clinically protective effect.

Matching provides a transparent descriptive baseline and an empirical-support warning system.  Most prespecified exposures produced summaries, and 14 of 15 successful original-cohort matching summaries shared the primary forest direction.  Failures identified cases in which no usable binary matching columns remained; some successful rows carried insufficient-pair warnings; and accepted Hamming distances varied.  Stable matching availability across original and downsampled populations supports the reproducibility of the availability pattern, but matching neither proves positivity nor supplies complete balance evidence.  A failed comparison is missing support, not a zero estimate, and a successful descriptive matched-pair outcome difference is not automatically an ATT or independent causal confirmation.

Sensitivity and permutation evidence adds layers rather than closure.  DML sensitivity outputs were saved-training direct, reconstructed, partial, or failed depending on the exposure, and reconstructed diagnostics are not identical to estimator-native calculations.  Benchmark-confounder comparisons supply a scale, not proof that omitted-variable bias is absent.  Treatment and outcome permutations provide disruption-based sanity checks for the DML estimators, but the archived procedure is not automatically a formal randomization test and does not establish identification.  The corresponding CausalPFN stages were intentionally skipped.  Finally, matching fields provide indirect support evidence, but dedicated propensity, common-support, and balance plots were not archived; positivity remains unestablished \cite{crump_et_al_2009_limited_overlap,cinelli_hazlett_2020_sensitivity,chernozhukov_et_al_2026_ovb_causal_ml}.

\subsection{Cross-Dataset Interpretation and Robustness}
\label{subsec:discussion:cross-dataset-robustness}

The original causal-analysis populations contained 26,845 MIMIC records and 7,993 PhysioNet records, so the MIMIC analysis population was approximately 3.36 times larger.  A larger analysis population can provide more information for nuisance-model fitting and modeled effect heterogeneity, and it may contribute to numerical stability.  It does not demonstrate better data quality, make the MIMIC findings more valid, or erase differences in era, variables, measurement systems, missingness, proxy rules, DAGs, or patient composition.  Predictive training and causal analysis also use distinct artifact contracts, so their populations should not be assumed identical merely because some row counts align.

Broad methodological replication across both datasets is a strength: the pipeline operated with the same overall sequence of representation, proxy prediction, aggregation, graph-guided adjustment, and estimator comparison.  However, similarly named proxy states are not automatically construct-equivalent.  MIMIC combines some domains that PhysioNet separates, and the rules and graphs are dataset specific.  The deliberate no-pooling policy is therefore essential.  Cross-dataset differences are evidence about portability of the workflow and sensitivity to context, not a pooled clinical effect or proof of biological inconsistency.

Outcome-based downsampling remained a contained robustness analysis.  It retained outcome-positive records and sampled outcome-negative records, thereby changing the empirical population and outcome prevalence.  Original and downsampled means do not automatically target the same population quantity.  Nevertheless, direction was preserved in 55 of 57 matched estimator--dataset--exposure comparisons, which indicates broad sign stability under a substantial sampling perturbation and suggests that many sign patterns were not solely driven by the original mortality imbalance.  Every paired mean changed numerically, and two PhysioNet comparisons changed sign.  The procedure is not formal transport or reweighting; magnitude changes cannot be read as stronger or weaker original-population effects, and directional preservation does not validate causal identification.

\subsection{Relationship to Prior Work and Methodological Implications}
\label{subsec:discussion:prior-work-implications}

The contribution lies in connecting layers that are often evaluated separately.  First, LLM-assisted elicitation converts unstructured clinical and causal reasoning into explicit proposals that can be selected, source-encoded, and audited.  Second, a shared proxy-state interface links irregular-time-series prediction and observational effect-estimation workflows.  Third, explicit separation of causal and split-aware predictive data contracts prevents silent identifier, split, and cohort conflation.  Fourth, source-coded DAGs make adjustment choices inspectable even when their scientific correctness remains an assumption.  Fifth, multi-estimator comparison exposes both broad directional agreement and important exceptions such as PhysioNet shock.  Matching, sensitivity, and permutation diagnostics reveal support and provenance limitations that a single effect table would conceal, while result manifests and validated source tables materially improve numerical traceability in a complex workflow.

These implications align with the wider emphasis on explicit intervention definitions and target-trial thinking in observational research \cite{hernan_robins_2016_target_trial,hernan_taubman_2008_well_defined_interventions}.  They also address difficulties identified in observational ICU causal analysis and machine-learning HTE research: defining the target question, controlling bias, validating heterogeneity, handling covariate shift, and distinguishing predictive variation from actionable treatment-effect variation \cite{smit_2023_causal_inference_icu_scoping_review,bica_2021_individualized_treatment_effects_ehr_ml,lipkovich_2024_modern_hte_methods,curth_2024_ml_individualized_treatment_effects,iwashyna_2015_hte_critical_care}.  The literature supplies methodological context; it does not validate this project's proxy labels, graphs, estimates, or local execution.  Accordingly, the framework should be described as an auditable research workflow rather than a generally validated clinical methodology.

\subsection{Clinical Meaning and Appropriate Use}
\label{subsec:discussion:clinical-meaning}

The proxy states summarize clinically inspired patterns and are analytically useful for organizing retrospective evidence.  They may support hypothesis generation and help identify proxy-state contrasts that merit targeted clinical and causal investigation.  High prediction performance against rule labels does not establish clinical accuracy, however, and a positive adjusted proxy-state contrast does not imply that clinicians should attempt to manipulate the proxy directly.  The system did not undergo prospective validation, external clinical validation, or clinical impact analysis.  Its causal quantities are not ready for bedside decision support, and no patient-level treatment recommendation follows from the results.

\paragraph{SRQ-7---principal limitations.}
The main limitations arise from proxy-label construct validity and measurement error; ill-defined or non-manipulable exposures; unvalidated DAG assumptions and possible unmeasured confounding; temporal ambiguity; incomplete empirical support; model dependence; outcome-class downsampling; dataset heterogeneity; incomplete final-run provenance; and the absence of clinical, fairness, and deployment validation.  The answer is therefore \emph{empirically supported with substantial unresolved boundaries}.  These limitations determine the strength of every preceding answer and motivate the validity analysis below.

\begin{longtable}{@{}L{0.10\textwidth}L{0.34\textwidth}L{0.18\textwidth}L{0.28\textwidth}@{}}
\caption{Evidence-based answers to the research questions.}
\label{tab:discussion-rq-answers}\\
\toprule
Question & Evidence-based answer & Strength of answer & Principal boundary \\
\midrule
\endfirsthead
\toprule
Question & Evidence-based answer & Strength of answer & Principal boundary \\
\midrule
\endhead
\bottomrule
\endfoot
Main RQ & An auditable workflow links LLM-assisted design, irregular measurements, deterministic proxy construction, deep prediction, algorithmic aggregation, project DAGs, and adjusted analyses across both datasets. & Implementation supported & Clinical and causal meaning remains assumption dependent. \\
SRQ-1 & Separate causal and split-aware predictive contracts are connected through normalized identifiers, routing, and explicit export schemas. & Implementation supported & Producing hashes, commands, and split lineage are incomplete. \\
SRQ-2 & Prompt, source, and generated artifacts trace selected LLM-assisted proxy, rule, and DAG designs at schema level. & Partially traceable; unresolved clinically & Human selection, line-level correspondence, chart adjudication, and graph review are incomplete. \\
SRQ-3 & The included model families completed the proxy-label task; the leading family differed by dataset. & Empirically supported with qualifications & No uncertainty or paired superiority tests; target validity and provenance are incomplete. \\
SRQ-4 & Exports are normalized to probability/binary pairs; final-run votes combine one rule source with four prediction sources. & Implementation supported & Voting is not clinical consensus and can preserve shared error. \\
SRQ-5 & Source-coded DAGs yield exposure-specific observed adjustment sets. & Conditional on assumptions & Graph correctness, temporal ordering, and identifiability are unresolved. \\
SRQ-6 & Primary, secondary, exploratory, matching, and diagnostic analyses show broad directional triangulation with explicit exceptions. & Empirically supported with qualifications & Agreement is not estimator equivalence or causal validation. \\
SRQ-7 & Construct, identification, support, modeling, transport, provenance, and clinical-use limitations materially bound all findings. & Partially supported & Several limitations require new clinical review, provenance, or validation evidence. \\
\end{longtable}

\section{Limitations and Threats to Validity}
\label{sec:discussion:limitations-threats-validity}
% Section ID: C11.2.
% Evidence map: see thesis-writing/planning/chapter_evidence_map.md.

\subsection{Construct and Measurement Validity}
\label{subsec:discussion:construct-measurement-validity}

The principal construct limitation is that clinically inspired rules approximate conditions but were not evaluated against chart-adjudicated diagnoses.  Available evidence consists of active rule code, binary artifact schemas, and limited descriptive validation summaries.  This makes the constructs transparent but does not establish their sensitivity, specificity, or equivalence across datasets.  Prediction can add classification error, and the mixed-source majority vote can preserve shared systematic error among models trained on the same rules.  Existing mitigations are deterministic definitions, source-code traceability, dataset-specific naming, and explicit proxy terminology; the residual risk remains high because clinical review and reference-standard validation are absent.

Binary exposure states also compress severity, duration, and timing.  A positive tag may combine clinically distinct trajectories, while a negative tag can reflect missing measurement rather than absence of a condition.  Proxy onset relative to covariates and mortality is not fully established, so exposure misclassification and temporal ambiguity can bias both prediction and adjusted analyses.  The processed outcome, \texttt{in\_hospital\_mortality}, is consistently required by the software, but its producing-artifact provenance is incomplete, and no competing-risk or post-discharge mortality analysis was performed.

The observation process is itself informative.  Measurement frequency may reflect clinical concern, resource use, or care pathways; models that encode masks or time gaps can exploit this signal without removing its causal implications.  Preprocessing and missingness-aware representation are therefore not equivalent to adjustment for the decision to measure.  This concern can affect proxy construction, predictive comparisons, and effect estimation simultaneously.

\subsection{Causal Identification and Estimand Validity}
\label{subsec:discussion:identification-estimand-validity}

These are the central constraints on causal interpretation.  Illness-state proxies are not necessarily well-defined interventions, so consistency and the meaning of contrasting exposure states may be difficult to specify.  No target-trial emulation was performed, and the pipeline does not implement a method for time-varying treatment--confounder feedback.  Exchangeability depends on a project-specified DAG, correct temporal ordering, and measured variables; topology alone cannot establish patient-record timing.  Some archived observed adjustment sets are empty, every checked identifiability field remains unresolved, and unmeasured or misspecified confounding remains possible.

Positivity was assumed rather than established.  Matching availability, failures, pair counts, and distances provide indirect evidence, but the archive lacks dedicated propensity distributions, common-support plots, and pre/post-match balance summaries.  Interference is also assumed, not tested, and repeated admissions or shared ICU processes may complicate unit-level independence.

Estimand language remains deliberately restricted.  The mean model-estimated CATE over the analyzed sample is an aggregate of model outputs, not automatically a population ATE.  The descriptive matched-pair outcome difference is not automatically an ATT.  Normalized CATE was omitted because scaling by the sample outcome rate is unstable across populations and does not create a risk ratio or another standard clinical estimand.  These boundaries mitigate overstatement but do not resolve the underlying intervention-definition and identification problems.

\subsection{Statistical, Modeling, and Support Limitations}
\label{subsec:discussion:statistical-modeling-support}

Many displayed main estimates lack uncertainty intervals, no formal multiple-comparison strategy was documented, and this thesis did not add significance tests.  Absence of an interval cannot be interpreted as either significance or insignificance.  Model selection, nuisance-model adequacy, regularization, hyperparameters, and cross-fitting behavior remain assumptions whose impact was not exhaustively tested.  CausalForestDML, LinearDML, and CausalPFN do not provide interchangeable uncertainty, and directional agreement can coexist with material magnitude and ordering differences.

Support evidence is incomplete and representation dependent.  Some matching runs failed, some successful matches required larger Hamming distances, and some carried insufficient-pair warnings.  Sensitivity outputs were partial or failed for several exposure families; reconstructed or fallback diagnostics have different status from estimator-native outputs.  Permutation aggregates are limited sanity checks rather than formal inferential tests.  CausalPFN lacks the corresponding archived downstream diagnostic family.  The multi-estimator and diagnostic hierarchy exposes these risks, but it cannot remove them.

\subsection{External Validity and Generalizability}
\label{subsec:discussion:external-validity}

MIMIC-III and PhysioNet differ in size, era, variables, measurement practices, and patient composition.  Their proxy ontologies and DAGs are dataset specific, and nominally similar states are not necessarily equivalent.  Numerical estimates were therefore not pooled.  Operation of the workflow on both sources is positive evidence of engineering and methodological portability, especially given the inclusion of several predictive and causal model families.  It is not evidence that the estimates transport to other hospitals, countries, care practices, EHR systems, or time periods.

No prospective validation, independent external clinical validation, or formal transport analysis was performed.  Subgroup performance and effect stability by age, gender, ICU type, or other characteristics were not evaluated as a fairness study.  Consequently, the analysis does not establish clinical generalizability or subgroup equity.

\subsection{Reproducibility and Provenance}
\label{subsec:discussion:reproducibility-provenance}

The validated result manifest, source tables, source paths, and checksums provide strong numerical-transcription support for the values reported in Chapter~\ref{chap:results}.  They also improve artifact traceability by distinguishing canonical sources, duplicates, and excluded artifacts.  This is not the same as computational rerun reproducibility.  The \texttt{final-results/} archive is ignored or untracked in the main repository; exact numbered causal configurations are unavailable; result-generating source versions and copy histories are incomplete; and the nested causal repository contains local modifications.

Predictive split provenance, checkpoint-to-export mapping, and copied-output lineage remain incomplete.  Raw and processed datasets are external, exact processed-pickle hashes and producing commands are unavailable, and some paths are absolute to the producing machine.  Final software, environment, and hardware records are also incomplete.  Thus the available records support archived numerical transcription and materially improve artifact traceability, while a clean-checkout rerun is not currently demonstrated.  Scientific reproducibility---replication of findings under independently reconstructed data, definitions, and analysis---is weaker still and requires new evidence rather than additional formatting of the archive.

\subsection{LLM-Assisted Design Layer}
\label{subsec:discussion:llm-provenance}

The LLM-assisted design layer contributed structured proposals for proxy concepts, binary decision rules, missingness considerations, and dataset-specific DAGs.  Potential advantages are the systematic elicitation of domain concepts, scalable generation of candidate proxy definitions and rules, parallel dataset-specific proposals, explicit prompt artifacts that improve design provenance, and conversion of otherwise unstructured reasoning into inspectable source-coded artifacts.  These advantages concern organization, traceability, and design productivity; the thesis did not compare them with clinician-only construction or measure whether LLM assistance improved clinical or causal correctness.

The main design risks include hallucinated or unsupported clinical assertions; sensitivity to prompt wording, supplied context, model version, and hosted-model updates; unknown system, browsing, and decoding settings; incomplete reproducibility of hosted-model behavior; and possible dependence on hidden training data.  Project selection can introduce anchoring and confirmation bias, especially when plausible generated explanations are mistaken for independent evidence.  The prompt record does not establish complete accepted/rejected decisions, and only high-level rather than line-by-line prompt-output-to-code traceability is available.

Validation gaps remain substantial.  There is no documented clinician-panel adjudication of every proxy definition or graph edge, no chart-level proxy validation, no comparison of LLM-assisted with clinician-only or hybrid construction, no replication with another LLM family, and no prompt-perturbation or model-version sensitivity experiment.  Error can therefore propagate from an unsupported design proposal into a deterministic label, from that label into a prediction and aggregate, and from a misspecified exposure or graph into adjustment and effect estimation.  Preserved prompts and explicit source-code authority make the chain inspectable but do not validate it.

Predictive agreement with LLM-assisted rule-derived labels validates neither the clinical construct nor the LLM-generated design.  Sensitivity and permutation diagnostics do not validate the proxy ontology or DAG.  No claim is made that an LLM provides authoritative clinical judgment, discovered the true causal graph, or can replace a clinician or causal expert.

\subsection{Ethical and Clinical-Deployment Considerations}
\label{subsec:discussion:ethics-deployment}

Misclassification can misrepresent a record's clinical condition, amplify measurement bias, and generate misleading exposure contrasts.  Groups with different testing intensity or access to measurements may be affected disproportionately.  Automation bias is therefore a direct risk: users could overread model probabilities as diagnoses, proxy-state rankings as clinical priorities, adjusted estimates as treatment recommendations, or estimator agreement as causal certainty.

No dedicated fairness audit was performed.  The presence of age, gender, ICU type, or other variables in an adjustment set does not constitute fairness validation, and subgroup predictive performance and effect stability remain untested.  The repository does not establish the project-specific institutional approval, consent or waiver procedure, data-use agreement, or governance determination.  These facts must be supplied from authoritative institutional records and must not be inferred from dataset access, de-identification, or repository contents.

The deployment boundary is unambiguous: \emph{the system is a retrospective research workflow and is not suitable for patient-level clinical deployment or treatment recommendation in its current form}.  No prospective safety, workflow, impact, or human-factors evaluation was performed.  This limitation is not cured by predictive accuracy, estimator agreement, or artifact traceability.

\begingroup
\scriptsize
\sloppy
\setlength{\tabcolsep}{2pt}
\begin{longtable}{@{}L{0.14\textwidth}L{0.24\textwidth}L{0.18\textwidth}L{0.19\textwidth}L{0.17\textwidth}@{}}
\caption{Study-specific limitations and threats to validity.}
\label{tab:discussion-limitations}\\
\toprule
Validity domain & Repository-specific limitation & Potential impact & Existing mitigation & Residual risk \\
\midrule
\endfirsthead
\toprule
Validity domain & Repository-specific limitation & Potential impact & Existing mitigation & Residual risk \\
\midrule
\endhead
\bottomrule
\endfoot
Proxy construct validity & Rule-defined states lack chart-adjudicated reference labels. & Construct error can distort prediction targets and exposure contrasts. & Explicit rules, proxy terminology, and source traceability. & High; clinical review and validation remain absent. \\
Prediction-label validity & Models learn rule-derived targets and voters may share error. & Performance can be high against an imperfect label source. & Separate rule, prediction, and vote terminology. & High; voting cannot remove correlated systematic error. \\
Exposure definition & Binary illness-state proxies are not necessarily manipulable interventions. & Consistency and causal meaning may be ill defined. & Exposure wording and no treatment recommendation. & High; intervention definition remains unresolved. \\
Temporal ordering & Proxy, covariate, and outcome timing is not established by graph topology. & Reverse ordering or post-exposure adjustment can bias estimates. & DAG descendant filtering and explicit caveats. & High; record-level timing needs validation. \\
DAG validity & Project graphs were not learned or empirically validated. & Misspecified arrows can select inadequate adjustment sets. & Source-coded transparency and prompt/code traceability. & High; expert review is incomplete. \\
Unmeasured confounding & Available variables may not block real-world backdoor paths. & Adjusted estimates may retain bias. & DML adjustment and layered sensitivity diagnostics. & High; sensitivity does not eliminate omitted-variable bias. \\
Overlap and support & Dedicated propensity and balance diagnostics are absent; matching failures occur. & Weakly supported contrasts may be unstable or nontransportable. & Pair, rate, distance, warning, and failure fields. & High; positivity is unestablished. \\
Estimator/model dependence & Estimators differ in structure, magnitude, ranking, and diagnostics. & Conclusions may depend on model form or tuning. & Prespecified hierarchy and multi-estimator triangulation. & Medium--high; agreement is mainly directional. \\
Uncertainty and comparisons & Intervals are unavailable for many displayed means; no multiplicity plan exists. & Precision and false-discovery risk cannot be fully assessed. & No new significance claims or tests. & High for exposure-specific inference. \\
Outcome downsampling & Sampling changes outcome prevalence and the empirical population. & Means no longer target an automatically identical population quantity. & Robustness-only role, no pooling, explicit sign-change report. & Medium--high; some conclusions are sampling sensitive. \\
Cross-dataset transport & Datasets differ in era, size, variables, proxies, DAGs, and populations. & Local patterns may not transport or be construct equivalent. & Dataset-specific reporting and no numerical pooling. & High; no external transport validation. \\
Missing result configurations & Numbered producing causal configs are unavailable. & Exact reruns and setting verification are blocked. & Run summaries and result manifest. & High for computational reproduction. \\
Predictive provenance & Split, checkpoint-to-export, and copy lineage are incomplete. & Final export and out-of-sample claims remain qualified. & Schema checks, paired logs, and wrappers. & High until per-export lineage is recovered. \\
Ignored result archive & \texttt{final-results/} is ignored or untracked. & Clean-checkout artifact recovery is not guaranteed. & Checksums, source packet, and canonical-source register. & Medium--high; external archival record is needed. \\
LLM-assisted design & Prompt/system settings, model sensitivity, accepted/rejected decisions, and line-level code traceability are incomplete. & Hallucination, anchoring, hidden-data dependence, and propagated rule or graph error are possible. & Preserved prompts, explicit source-code authority, and conservative claim boundaries. & High until blinded clinical review, replication, and sensitivity studies are completed. \\
Clinical validation & No prospective, chart-review, or external clinical validation was performed. & Clinical accuracy and utility are unknown. & Retrospective, hypothesis-generating use boundary. & High; bedside use is inappropriate. \\
Fairness and deployment & No subgroup fairness, safety, impact, or human-factors audit was performed. & Unequal errors and automation harm may go undetected. & Explicit non-deployment boundary. & High; dedicated evaluation is required. \\
\end{longtable}
\endgroup
