\chapter{Conclusions and Future Work}
\label{chap:conclusions-future-work}

\section{Summary of the Study}
\label{sec:conclusions-future-work:summary-of-the-study}
% Section ID: C12.1.
% Evidence map: see thesis-writing/planning/chapter_evidence_map.md.

This thesis addressed the challenge of connecting irregular ICU time-series data to transparent retrospective analysis without concealing the transformations required along the way.  The motivating problem is not only that clinical measurements are sparse, asynchronous, and shaped by the measurement process, but also that design assumptions, prediction, and observational analysis can become disconnected when their labels, identifiers, preprocessing contracts, and interpretation boundaries are not explicit.  The study therefore developed and evaluated an evidence-tracked workflow across PhysioNet 2012 and MIMIC-III that combines LLM-assisted domain-knowledge construction, deterministic rule-derived proxy labeling, deep proxy-state prediction, normalized and mixed-source aggregated proxy labels, LLM-assisted project-specified DAGs, matching, heterogeneous-effect estimation, and available diagnostics.

The resulting framework is a retrospective research workflow.  Its proxy states are rule-derived analytical constructs rather than verified clinical conditions, and its adjusted estimates remain conditional on the measurement, graph, intervention-definition, temporal-ordering, support, and modeling assumptions stated throughout the thesis.  This distinction is central to the contribution: the workflow makes its handoffs, evidence sources, and unresolved requirements visible instead of presenting an apparently seamless output as a clinical fact.

\section{Main Findings and Contributions}
\label{sec:conclusions-future-work:main-findings-contributions}

The strongest contribution is the integrated, evidence-tracked connection between LLM-assisted proxy and DAG design, irregular-time-series representation, deterministic proxy-label construction, deep predictive modeling, algorithmic aggregation, and downstream observational analysis.  The shared proxy-state interface provides a traceable route from selected source-code rules to rule-derived prediction targets, binary predicted voter inputs, a mixed-source vote combining one rule-derived source with four prediction sources, and dataset-specific exposure analyses.  Separate causal and split-aware predictive contracts further make identifier, cohort, and export boundaries inspectable.  Checked result tables, manifests, checksums, source packets, and explicit deferred issues complement this workflow by improving numerical traceability, even though they do not yet provide a complete clean-checkout reproduction record.

The empirical predictive comparison covered STraTS, GRU, GRU-D, TCN, and SAnD.  The leading family differed by dataset: STraTS led the archived proxy-label test metrics in MIMIC-III, whereas GRU-D led them in PhysioNet (Chapter~\ref{chap:results}, Section~\ref{sec:results:predictive-performance}).  This result supports a dataset-specific view of representation choice.  It does not establish a universal model ordering, statistical superiority, or clinical validity of the rule-derived targets.

In the primary original-cohort analysis, CausalForestDML produced positive mean model-estimated CATE summaries for all nine MIMIC-III exposures and for nine of ten PhysioNet exposures; the PhysioNet shock summary was negative (Chapter~\ref{chap:results}).  These estimates document systematic variation in the constructed proxy-state analyses after the implemented adjustment procedure; they do not establish that a proxy state caused mortality or that modifying it would change an outcome.  Matching supplied a descriptive empirical-support baseline, not independent confirmation, and its failures and weak-support flags remain part of the evidence.

The comparison between CausalForestDML and LinearDML adds useful but bounded triangulation: their mean summaries had the same direction in all 19 original-cohort dataset--exposure comparisons, while magnitudes and within-dataset rankings still differed (Chapter~\ref{chap:results}).  Thus agreement is informative about the observed sign pattern but not a guarantee of estimator correctness, equal uncertainty, or valid identification.  CausalPFN also agreed in direction with both DML estimators in 18 of 19 comparisons, with PhysioNet shock as the exception (Chapter~\ref{chap:results}).  This is an exploratory complementary finding only: its primary method source is unresolved in the approved corpus, and its sensitivity and permutation stages were intentionally not part of the archived diagnostic workflow.

Together, these results support the thesis's methodological objective: LLM-assisted knowledge construction, deterministic proxy labeling, deep proxy-state prediction, and DAG-guided causal-effect estimation can be connected while preserving dataset-specific definitions, estimator hierarchy, and evidence boundaries.  The contribution is the transparent integration and qualification of these components, not proof that the LLM-generated proposals, proxy constructs, graphs, or effect estimates are clinically correct.

\section{Limitations of the Conclusions}
\label{sec:conclusions-future-work:limitations-of-the-conclusions}

The conclusions are constrained first by construct validity.  The proxy states originate in LLM-assisted, project-selected rule designs encoded as deterministic source; they are not chart-adjudicated labels, and the two datasets differ in measured variables, proxy ontologies, and timing.  Unsupported design assertions can enter the rules, prediction can add error, and mixed-source majority voting can preserve shared rule and measurement error; a binary tag can compress distinct trajectories, and a negative tag can reflect unavailable measurement rather than absence of a condition.  Clinical review, reference-standard evaluation, complete prompt-to-code traceability, and exact support for several rule families remain incomplete.

Second, the observational analyses do not remove identification uncertainty.  Illness-state proxy exposures are not necessarily well-defined interventions, the LLM-assisted project-specified DAGs are encoded assumptions rather than empirically established structures, temporal ordering is incomplete, and unmeasured confounding can remain.  An LLM-assisted graph is not intrinsically more correct than a manually proposed graph.  Matching rates, distances, failures, sensitivity analyses, and permutation checks qualify particular risks but do not establish global overlap, balance, exchangeability, proxy or graph validity, or causal certainty.  Estimator agreement must therefore be read alongside estimator disagreement, support gaps, incomplete uncertainty reporting, and the different diagnostic coverage of CausalPFN.

Finally, the present evidence does not establish external clinical utility or evaluate the LLM as a design method against alternatives.  The two datasets support application of the workflow in distinct settings but do not justify numerical pooling or transport claims.  No external or prospective clinical validation, subgroup fairness analysis, safety study, impact evaluation, patient-level deployment evaluation, clinician-only comparison, alternative-LLM replication, or prompt-sensitivity study was performed.  Authoritative institutional ethics and data-governance documentation also remains required.  Reproducibility is incomplete: raw and processed data artifacts, exact numbered causal configurations, predictive split and checkpoint lineage, producing source versions, archive-copy history, and full LLM prompt-run and human-review records remain unresolved.

\section{Future Work}
\label{sec:conclusions-future-work:future-work}

The first priority is clinical and construct validation.  Future work should freeze the proxy rules and conduct blinded clinician review of every proxy definition and graph edge, with multiple reviewers and reported inter-reviewer agreement.  Rule-derived and predicted states should be compared with chart-adjudicated reference labels, including dataset-specific assessment of timing, missingness, thresholds, and differences among rule-derived labels, individual predicted labels, and mixed-source aggregated proxy labels.  A controlled comparison of LLM-assisted, clinician-only, and hybrid rule and DAG construction would test whether the design layer adds value rather than merely plausibility.

The second priority is a stronger observational and measurement design.  A future study should specify a target-trial-aligned question with eligibility, time zero, intervention strategies, follow-up, outcome, estimand, and an explicit plan for time-varying confounding.  It should add prespecified overlap and balance diagnostics, support-aware estimands, uncertainty and multiplicity procedures, and richer sensitivity analyses.  Rule ablation and threshold-sensitivity experiments, formal label-noise modeling, and explicit measurement-error analyses should quantify how proxy construction, prediction, and aggregation affect downstream estimates.  DAG review should proceed edge by edge and include alternative-graph and adjustment-set analyses rather than treating one encoded graph as fixed truth.

The third priority is reproducibility and external evaluation.  Signed per-run manifests should link raw-data access records, processed-artifact hashes, split identifiers, checkpoints, commands, resolved configurations, software environments, source versions, and archived outputs.  LLM design manifests should record complete prompts, system and decoding settings, model versions, follow-up interactions, reviewer decisions, and prompt-output-to-code mappings.  Prompt perturbations, model-version sensitivity studies, and replication across multiple LLM families should be performed.  Frozen definitions should then be evaluated on temporal and external cohorts and in prospective studies of clinically justified subgroups before any safety or impact study is considered.  CausalPFN should remain exploratory until its primary source and implementation version are verified and an appropriate diagnostic and uncertainty plan is available.  Authoritative ethics and data-governance wording should also be completed before prospective or deployment-related evaluation.

\section{Closing Perspective}
\label{sec:conclusions-future-work:closing-perspective}

The thesis shows the value of treating integration itself as an object of evidence.  LLM-assisted knowledge construction, irregular ICU records, deterministic proxy labels, deep predictive outputs, graph-guided adjustment, and model-estimated contrasts can be connected in a transparent workflow, but none of those handoffs removes the need for clinical review, careful observational design, diagnostic scrutiny, or reproducible provenance.  An LLM cannot replace a clinician or causal expert on the evidence presented here.  The appropriate conclusion is therefore measured: transparent integration can make retrospective evidence more inspectable and more useful for hypothesis generation, while the boundaries that prevent stronger clinical or causal claims remain visible and actionable for future research.
